The FAIRification of research in real-world evidence: A practical introduction to reproducible analytic workflows using Git and R

### 2023-07-04 manuscript/manuscript.qmd
added key points and made minor modifications here and there

[**Word count:**]{.underline} 3,239 words

@@

@@

Real-world evidence (RWE) studies that make secondary use of routinely collected health data captured in electronic health records and claims are increasingly being used to inform drug development, regulatory and coverage decisions, as well as clinical practice.[@fdaRWE2018; @eichler2015; @cioms2023; @makady2017] Considering the impact that RWE studies can have on patient healthcare, transparent and reproducible conduct of RWE studies is critically important. While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python. Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community. As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.

@@

At this step, we now can populate the repository and start working on study-relevant documents and analyses. To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure). To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file. This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added aftwerwards by creating a text file with the name `README.md` in the root directory of the repository.

To speed up these initial steps, we provide a pre-populated template of such a structured study repository based on the HARPER protocol[@wang2022] template under <https://gitlab-scm.partners.org/drugepi/harper>. HARPER is a harmonized study protocol template endorsed by the International Society of Pharmacoepidemiology and the International Society of Pharmacoeconomics and Outcomes Research. The pre-populated structured study repository template is publicly available and can be `forked`[@GitFork] to initiate a new project (see instructions in the template repository).

@@

This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems. However, most of the times researchers use `Git` for the purpose of collaborating on a project as depicted in @fig-collaborate.

@@

One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project. Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>). To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step to necessary is running the following command in R.

@@

The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming. A better solution to this is the use of relative paths. A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows. The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory). Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any user re-running this script.

```{r}
library(here)
icd_codes <- read.csv(here("data", "icd_codes.csv"))
```

@@

Moreover, when working with remote repositories, it is of utmost importance to ensure that sensitive information (such as passwords or patient-level data) are not accidentally shared. To prevent this, we recommend specifying such files in the `.gitignore` file before making the first commit or, more preferably, not storing them in the project directory in the first place.

@@

Regulators[@fdaRWE2021], funding agencies[@NIH2023;@PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies. This manuscript addresses this timely topic and will hopefully motivate and enable new and experienced pharmacoepidemiologists to integrate VCS workflows and code sharing into their daily work.

@@

```{r}
#| label: fig-GitHub-gui_changes
#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures. This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom)."
#| eval: true
#| echo: false
#| out-width: 105%

>>>>

[**Word count:**]{.underline} 3,336 words

@@

::: {.callout-note icon="false"}
## KEY POINTS

-   Transparency and reproducibility are major prerequisites for conducting meaningful real-world evidence (RWE) studies that are fit for decision-making.

-   Regulators, funding agencies and HTA bodies have increasing expectations regarding the provenance, audit trial and sharing of study documents including analytic code as a prerequisite for impactful and credible RWE studies.

-   Version control systems (VCS) for analytic code like Git, GitHub and GitLab have unique advantages for making analytic code and other relevant RWE study documents Findable, Accessible, Interoperable and Reproducible (FAIR).

-   In this manuscript, we discuss and demonstrate how VCS can be used to track changes over the course of a RWE project, collaborate, disseminate and archive analytic code and provide a technical step-by-step tutorial for the practical implementation of reproducible workflows using Git.

-   Given the trend towards open source tools for clinical trial reporting and regulatory submissions, this tutorial addresses a timely topic and aims to encourage new and experienced pharmacoepidemiologists to integrate VCS in their daily work and embrace the advantages of FAIR analytic code sharing.
:::

@@

Real-world evidence (RWE) studies that make secondary use of routinely collected health data captured in electronic health records and claims are increasingly being used to inform drug development, regulatory and coverage decisions, as well as clinical practice.[@fdaRWE2018; @eichler2015; @cioms2023; @makady2017] Considering the impact that RWE studies can have on patient healthcare, transparent and reproducible conduct of RWE studies is critically important.

While many advances have been made in the documentation and reporting of study protocols and results[@wang2021; @schneeweiss2019; @berger2017; @wang2022], guidance on transparency regarding the actual implementation and analytic steps is still significantly lacking, especially when analyses are implemented via line coding in statistical programming languages like R[@RCore2023], SAS or Python. Although in other quantitative disciplines such as computational biology[@almugbel2017; @mammoliti2021; @russell2018; @perez-riverol2016] or health informatics[@bakken2019], the sharing of code is rather the rule than the exception, this practice is not as common in the (pharmaco-) epidemiological community. As open source tools like R packages and Shiny dashboards[@pharmaverse2023] become increasingly used for clinical trial reporting and regulatory submissions[@positPharma], the adoption of version control and code sharing practices is a crucial and timely topic for the field of RWE.

@@

At this step, we now can populate the repository and start working on study-relevant documents and analyses. To enhance clarity, it is recommended to use a logical structure and an informative nomenclature for sub-directories and file names (a minimal example is illustrated in @fig-structure). To document the structure and location of files, we can include this, and other critical information about the project, in the `README.md` file. This file can be auto-generated when a project is initialized through a remote repository (see prior step in @sec-gui-initialization-and-git-clone) or can be manually added afterwards by creating a text file with the name `README.md` in the root directory of the repository.

To speed up these initial steps, we provide a pre-populated template of such a structured study repository based on the HARPER protocol[@wang2022] template under <https://gitlab-scm.partners.org/drugepi/harper>. HARPER is a harmonized study protocol template endorsed by the International Society of Pharmacoepidemiology and the International Society of Pharmacoeconomics and Outcomes Research. The pre-populated structured study repository template is publicly available and can be `forked`[@GitFork] to initiate a new project (see instructions in the template repository under the provided link).

@@

This step is not needed if the intention of a researcher is just to keep track of local changes and there is no wish to share code, collaborate with others or work across different systems. However, most of the times researchers use `Git` particularly for the purpose of collaborating on a project as depicted in @fig-collaborate.

@@

One way to achieve this is through self-contained computing environments via Docker images which are isolated and standardized environments which include everything needed to run and deploy applications and code regardless of a user's local operating system.[@boettiger2017a; @nüst2020] Another more lightweight, yet effective way is by using the `renv` R package[@renv], which helps creating **re**producible **env**ironments and manages R package dependencies for a given project. Through a lockfile, the `renv` package keeps track of the underlying version of R that is being used as well as all R packages, their versions and the installation source (e.g., <https://cran.r-project.org/>). To synchronize the R environment with that of a collaborator, e.g., after cloning a repository, the only step necessary is running the following command in R.

@@

The use of absolute paths will make it impossible for any collaborator to run the code without having to manually adapt every single path throughout a repository which is tedious and time consuming. A better solution to this is the use of relative paths. A popular R package to implement this is the `here` package[@here] which enables file referencing in project-oriented workflows. The main function "`here"` creates paths relative to the top-level of the project directory (in our example the `rwe_study` directory). Applied to our example project, the following command would automatically reference the correct path and could be used without modification by any collaborator re-running this script.

```{r}
library(here)
icd_codes <- read.csv(here("protocol", "icd_codes.csv"))
```

@@

Moreover, when working with remote repositories, it is of utmost importance to ensure that sensitive information (such as passwords or patient-level data) are not accidentally shared. To prevent this, we recommend specifying such files in the `.gitignore` file before making the very first commit or, more preferably, not storing them in the project directory in the first place.

@@

Regulators[@fdaRWE2021], funding agencies[@NIH2023;@PCORI2023] and HTA bodies[@kent2021] have increasing expectations regarding the provenance, sharing and audit trial of study documents including analytic code as a prerequisite for impactful and credible RWE studies. Given the trend towards open source tools for clinical trial reporting and regulatory submissions, this tutorial addresses a timely topic and hopefully encourages new and experienced pharmacoepidemiologists to integrate VCS in their daily work and embrace the advantages of FAIR analytic code sharing.

@@

```{r}
#| label: fig-GitHub-gui_changes
#| fig-cap: "Remote repositories like GitHub provide tools to visually track changes made to analytical code and corresponding output such as figures. This commit history illustrates details about the commit (e.g., commit SHA [simple hasing algorithm] and message), a side-by-side comparison of the previous (upper left) and modified (upper right) version of the figure and the corresponding changes made to the R code (bottom; line 32 in the code script was added)."
#| eval: true
#| echo: false
#| out-width: 105%
